---
title: "Fluconazole data analysis"
author: "R. Cunning"
date: "2024-02-15"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
# Load libraries
library(gam)
library(tidyverse)
library(lubridate)
library(janitor)

# Create custom ggplot theme
theme_custom <- function() {
  theme_bw(base_size = 10, base_family = "Arial") %+replace%
    theme(
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(), 
      panel.background = element_blank(),
      panel.border = element_rect(color = "black", fill = NA),
      legend.background = element_rect(fill = NA, colour = NA),
      axis.title = element_text(size = 7),
      axis.text = element_text(size = 5)
    )
}

# Function to pivot PAM data to long form with column for AOI
ipam_convert <- function(data) {
  data %>% select_if(~ !any(is.na(.))) %>%
    pivot_longer(cols = starts_with("f") | starts_with("y")) %>%
    separate(name, into = c("var", "aoi"), sep = "(?<=[A-Za-z_])(?=[0-9])")
}
```

```{r}
# Import PAM data
# List files with exported PAM data
pamfiles <- list.files(path = "data", pattern = "*.csv", full.names = T)

# Import data from each file
pam1 <- pamfiles %>%
  map_dfr(read_delim, delim = ";", .id = "file_id") %>%
  janitor::clean_names() %>%
  mutate(file_id = basename(pamfiles[as.numeric(file_id)]),
         date = as_date(date, format = "%d.%m.%y"))

# For files that have multiple sat pulses -- keep the last one only
pam1 <- pam1 %>%
  group_by(file_id, date) %>%
  filter(no == max(no)) %>%
  ungroup()

# For each source file, convert to long form data with F, FM, and YII for each AOI
pam1 <- pam1 %>%
  nest(-file_id, -date) %>%
  mutate(data2 = map(data, ipam_convert)) %>%
  unnest(data2) %>%
  group_by(file_id, date) %>%
  select(file_id, date, time, aoi, var, value) %>%
  mutate(aoi=as.numeric(aoi))

# Import metadata: order in which racks PAMmed, and order of genets on racks
pammd <- readxl::read_xlsx("Coral project part 2 data file.xlsx") %>%
  mutate(date = as_date(date)) %>%
  drop_na()

pammd<-pammd %>% 
  crossing(aoi=c(1,2,3)) %>%
  mutate(aoi=((position-1)*3)+aoi)

# Join PAM data with rack order information (which PAM file corresponds to which rack of corals)
pam <- full_join(pam1, pammd) %>%
  group_by(file_id, date) 

# Change treatment to a categorical variable
pam <- pam %>%
  mutate(treatment = factor(treatment, levels = c("0", "1", "5.3")))

# Check if any data and metadata did not match...(should be empty)
anti_join(pam1, pammd)

# Subset just the Fv/Fm data (YII)
yii <- pam %>%
  filter(var=="y_ii_")

# Write processed Fv/Fm data to file
write_csv(yii,"output/fvfm_processed.csv")
```

# Bryopsis Fv/Fm

```{r, fig.width = 10, fig.height = 6}
# Create subset of data for just Bryopsis
bry <- yii %>%
  filter(species=="Bryopsis") %>%
  mutate(area = case_when(aoi == 1 ~ "root",
                          aoi == 2 ~ "middle",
                          aoi == 3 ~ "edge"),
         area = factor(area, levels = c("root", "middle", "edge")))

# Filter out dead Bryopsis data
bryf <- bry %>%
  filter(!(tank == 4 & date > "2024-01-25"),
         !(tank == 7 & date > "2024-01-23"),
         !(tank == 2 & date > "2024-01-25"),
         !(tank == 8 & date > "2024-01-20"))

# Plot all Bryopsis fv/fm data
bryfig1 <- ggplot(bry, aes(x = date, y = value, group = area, color = area)) +
  geom_point() +
  geom_line() +
  geom_vline(aes(xintercept = as_date("2024-01-03")), lty = 2) +
  facet_wrap(~ treatment + tank, nrow = 3,
             labeller = labeller(.rows = label_both)) +
  scale_x_date(breaks = unique(bry$date)) +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(filename = "output/bryfig1.png", plot = bryfig1, 
       width = 120, height = 160, units = "mm")

bryfig1 <- bryfig1 +
  labs(title = "Fv/Fm for each area of each Bryopsis tuft over time (all raw data)")

filtered_bryfig1 <- bryfig1 %+% bryf +
  labs(title = "Filtered out 'dead' Bryopsis")

cowplot::plot_grid(bryfig1, filtered_bryfig1)

# Use filtered data for analysis
bry <- bryf
```


```{r, fig.height = 4, fig.width = 5}
# Linear model for fv/fm in Bryopsis at end of experiment 
library(lme4)
# Fit linear mixed model with treatment and date as fixed predictors, and tank as random
mod <- lmer(value ~ treatment * factor(date) + (1|tank/area), data = bry)

# Calculate the "marginal means" of each treatment on each date
library(emmeans)
all <- emmeans(mod, specs = "treatment", by = "date", level = 0.87)
res <- as_tibble(all)

# Plot means for each treatment over time
# Plot overall mean for each treatment over time
res %>%
  ggplot(aes(x = date, y = emmean, group = treatment, color = treatment)) +
  geom_point() + 
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.3) +
  geom_line() +
  geom_vline(aes(xintercept = as_date("2024-01-03")), lty = 2) +
  scale_x_date(breaks = unique(bry$date)) +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Average Fv/Fm of Bryopsis (all areas) in each treatment over time")

# Test for differences between treatments on each date
pwtests <- as_tibble(contrast(all, "pairwise"))

# When do we see statistical differences between treatments?
sig <- pwtests %>% filter(p.value < 0.05)
knitr::kable(sig)
```
When looking at overall average Fv/Fm for Bryopsis, there are no significant treatment effects, and confidence intervals are very wide... Let's look at just the edges of Bryopsis to see if anything clearer emerges.

### Just the edges of Bryopsis
```{r, fig.height = 4, fig.width = 5}
# Fit linear mixed model with treatment and date as fixed predictors, and tank as random
edges <- lmer(value ~ treatment * factor(date) + (1|tank), 
             data = filter(bry, area == "edge"))
# Means at each individual date
alledges <- emmeans(edges, specs = "treatment", by = "date", level = 0.87)

# Statistically compare these marginal means through pairwise contrasts
resedges <- as_tibble(contrast(alledges, "pairwise"))

# When do we see significant differences by treatment in the edges?
sigedges <- resedges %>%
  filter(p.value < 0.05)

as_tibble(alledges) %>%
  ggplot(aes(x = date, y = emmean, color = treatment, value = treatment)) +
  geom_point() + 
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.3) +
  geom_line() +
  geom_vline(aes(xintercept = as_date("2024-01-03")), lty = 2) +
  scale_x_date(breaks = unique(bry$date)) +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Average Fv/Fm of Bryopsis edges in each treatment over time")

knitr::kable(sigedges)
```

When looking at just the Bryopsis edges, both Treatment 1 and Treatment 5.3 were significantly lower than the control on January 18th. After January 18th, the increases seen in the treated Bryopsis probably reflect Fv/Fm signal of fouling algae.

The 1 mg/ml and 5.3 mg/ml treatments also don't look that different from one another, so we could combine these into "treated" vs. "untreated" (controls) to increase statistical power...

```{r, fig.height = 4, fig.width = 5}
#### Compare 2 control tanks vs. 4 "treated" tanks
bry2 <- bry %>%
  mutate(trt2 = case_when(as.numeric(as.character(treatment)) == 0 ~ "control",
                          as.numeric(as.character(treatment)) > 0  ~ "treated"),
         trt2 = factor(trt2))
# Fit linear mixed model with treatment and date as fixed predictors, and tank as random
mod<-lmer(value~trt2*factor(date)+(1|tank),data=filter(bry2, area == "edge"))
# Means at each individual date
alltrt2 <- emmeans(mod, specs = "trt2", by = "date", level = 0.87)
# Statistically compare these marginal means through pairwise contrasts
restrt2 <- contrast(alltrt2, "pairwise")
# there are differences between control and 4xtreated tanks, when looking at Bryopsis tips only

as_tibble(alltrt2) %>%
  ggplot(aes(x = date, y = emmean, group = trt2, color = trt2)) +
  geom_point() + 
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.3) +
  geom_line() +
  geom_vline(aes(xintercept = as_date("2024-01-03")), lty = 2) +
  scale_x_date(breaks = unique(bry$date)) +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Average Fv/Fm of all treated vs. control Bryopsis edges over time")

sigtrt2 <- as_tibble(restrt2) %>% 
  filter(p.value < 0.05)

# When do we see significant differences between control and treated?
knitr::kable(sigtrt2)
```

Combining the treated tanks together, we see that Bryopsis Fv/Fm (edges) is significantly lower in treated tanks vs. control on Jan. 13th, 16th, 18th, and 20th.


# Coral Fv/Fm
```{r, fig.width = 10, fig.height = 10}
# Create subset of data for just coral
coral <- yii %>%
  filter(species != "Bryopsis")


# Plot all coral fv/fm data
coralplot <- ggplot(coral, aes(x = date, y = value)) +
  geom_point() +
  facet_wrap(~ treatment + tank + species + fragment, nrow = 3,
             labeller = labeller(.rows = label_both)) +
  scale_x_date(breaks = unique(bry$date)) +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Identify outliers by smooth function
mod <- gam(value ~ s(date, spar = 0) * treatment * species * tank:fragment, data = coral)

# Plot data with outliers shown
coral <- coral %>%
  ungroup() %>%
  mutate(resid = residuals(mod),
         fit= fitted(mod))

ggplot(coral, aes(x = date, y = value)) +
  geom_point(aes(color = abs(resid) > 0.06)) +
  geom_line(aes(y = fit)) +
  facet_wrap(~ treatment + tank + species + fragment, nrow = 3,
             labeller = labeller(.rows = label_both)) +
  scale_x_date(breaks = unique(bry$date)) +
  scale_color_manual(values = c("black", "red")) +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Filter outliers
coral <- coral %>%
  filter(abs(resid) < 0.06)
```

```{r}
mod <- lmer(value ~ factor(date) * treatment * species + (1|tank/species/fragment),
            data = coral)

emm <- emmeans(mod, specs = "treatment", by = c("species", "date"), level = 0.87)

as_tibble(emm) %>%
  ggplot(aes(x = date, y = emmean, color = treatment, group = treatment)) +
  geom_point() +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.5) +
  geom_line() +
  scale_x_date(breaks = unique(bry$date)) +
  facet_wrap(~species) +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_vline(aes(xintercept = as_date("2024-01-03")), lty = 2)

res <- as_tibble(contrast(emm, "pairwise"))
res %>%
  filter(p.value < 0.05) %>%
  arrange(species, date)
```

Acropora corals showed a ~10% decline in Fv/Fm over time, but this occurred in all treatments and the control, so there is no effect of treatment on Fv/Fm in Acropora.

Seriatopora showed an initial decline in Fv/Fm that was larger in treated corals, such that treated corals. Treatment 1 mg/ml was significantly lower than control on 1/6, and Treatment 5.3 mg/ml was significantly lower than control on 1/13, 1/16, and 1/25 (I think the 1/25 readings might be erroneous). Fv/Fm recovers in the treated corals by the end of the experiment and is not different than the controls.

Stylophora also showed an initial decrease that also occurred in the controls. However, the treated corals are significantly lower than the controls during the experiment, on 1/6 and 1/16 (Treatment 1 is lower), and 1/20 and 1/25 (Treatment 5.3 is lower, but again I think that 1/25 readings seem off). Values again recover by the end of the experiment and are not different between treated and control corals.

Overall, there is some indication that fluconazole may temporarily reduce Fv/Fm in in Seriatopora and Stylophora, but these reductions were only around 5% relative to control values, and disappeared after ~2 weeks.




